---
title: "STAT 184 Final Project: Matt Bubb and Jacob Adams"
output: html_notebook
---
## Research idea: Expondential Growth and the Novel Coronavirus

#### Our final project is based around the concept of exponential growth when applied to the Novel Coronavirus (COVID-19). Although our main idea is not surrounded by a question, we have explored many sub-questions related to the topic in which we can draw trivial commonalities between the Coronavirus and exponential growth. Our research topics are listed below.  
<br/>

- Can we safely conclude that the cases worldwide are indeed spreading exponentially?
- If the total growth is exponential, which countries have had the highest growth rate in terms of total cases? 
- What is the average length of days it takes for the total number of COVID-19 cases to double?

<br/> 


```{r}
rm(list = ls())

library(DataComputing)
library(lubridate)
data1 <- read.csv("data1Corona.csv")
data2 <- read.csv("full_data.csv")
test <- data2 %>% group_by(date) %>%summarise(newCases = sum(new_cases),totalDeaths = sum(total_deaths) ,newDeaths = sum(new_deaths), totalCases = sum(total_cases)) %>% arrange(desc(date)) 
test$date = as.character.Date(levels(test$date))[test$date]
graph <- test%>%mutate(daysSinceDecember312019 = as.numeric(floor((difftime(test$date, min(test$date))/86400 )) ))
```





```{r}
ggplot(data = graph, aes(daysSinceDecember312019)) +  geom_point(aes(y = newCases, color = "New Cases")) + geom_point(aes(y = newDeaths, color = "New Deaths")) +geom_point(aes(y = totalDeaths, color = "Total Deaths")) +geom_point(aes(y = totalCases, color = "Total Cases"))
```

```{r}
ggplot(data = graph, aes(daysSinceDecember312019)) +  geom_point(aes(y = newCases, color = "New Cases")) + geom_point(aes(y = newDeaths, color = "New Deaths")) +geom_point(aes(y = totalDeaths, color = "Total Deaths")) 
```

```{r}
ggplot(data = graph, aes(daysSinceDecember312019))  + geom_point(aes(y = newDeaths, color = "New Deaths")) +geom_point(aes(y = totalDeaths, color = "Total Deaths")) 

```



Up next, we are interested to see how often the total number of cases around the world doubles. As the most common form of exponential growth $f(x) = 2^x$, grows at an incredible rate and in the case of the Coronavirus, the number of cases could grow incredibly fast. Below, I have have code below showing our calcualtions. 


```{r}




totalCasesColumn <- rev(test$totalCases)
total = 1
countDays = 0
doubleList = 0
n = 0
for (i in seq_along(totalCasesColumn)){
  countDays = countDays + 1
  checkForDouble = totalCasesColumn[i]
  if(checkForDouble > total * 2  ){
    total = checkForDouble
    doubleList = doubleList + countDays
    n = n + 1
    countDays = 0
  }
}

print(paste("On average the amount of days for COVID-19 cases to double is:", floor(doubleList / n)))

```

